Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-10-05 paper/paper_draft.qmd
Updates to PDA/carbonate budget results sections

Partial Dependence Analysis revealed reef median depth, carrying capacity and mean time
series DHW level have the highest feature importance scores, accounting for 29%, 15% and 15%
predictive importance respectively (@fig-ts-cluster-pdp; Panel A). Area weighted incoming
connectivity, outgoing connectivity strength, source-to-sink connectivity ratio and larval
retention probability account for \~12%, 11%, 10% and 8% respectively.

@@

![Feature importance scores (Panel A) of the Random Forest classifier, and Partial
Dependence Plots (PDPs) for reef characteristics of interest indicating the probability of a
reef being classified to one of the three cover clusters (Panel B). Partial Dependence Plots
for bioregions were excluded as they were found to be uninformative. PDPs indicate the
probability of a reef from the test dataset being clustered into each cover cluster
depending on the predictor value on the x axis, when adjusted for values of other
predictors. Green, orange and blue lines indicate cluster assignment probability for Low,
Medium and High clusters respectively. Transparent grey points above probability lines
represent the predictor values present in the test dataset, providing a rough indication of
predictor value distribution across
reefs.](../figs/ts_cluster_rf_feature_analysis.png){#fig-ts-cluster-pdp fig-align="center"
width="14.82cm"}

First-order partial dependence analysis showed minimal influence of incoming, outgoing and
source-to-sink ratio connectivity metrics on cluster assignment probabilities. Two-way
analyses revealed little interaction between depth and incoming and outgoing connectivity
metrics for reef clustering probabilities (@fig-ts-cluster-pdp-twoway-incoming;
@sfig-ts-cluster-pdp-twoway-outgoing). Reefs with a median depth around 15m and high
incoming connectivity may benefit slightly, with higher probability of High cover cluster
assignment compared to 15m - depth reefs with lower connectivity.

@@

Individual feature analysis for carrying capacity indicates that its influence on cluster
assignment is concentrated at the low range of predictor values (@fig-ts-cluster-pdp). This
aligns with the right-skewed distribution of carrying capacity across reefs for these three
predictor variables. Beyond this low-end range, the partial dependence becomes near
constant, indicating that outlier values for these factors are not influential. Assessing
random forest prediction accuracy across bioregions, reveals increasing performance as
bioregion mean reef depth increases (@sfig-rf-predictor-perf). This highlights that
unexplained variation in the test dataset (33% of variation) may be concentrated in shallow
bioregions where the available model predictors do not capture differences between clustered
reefs.

@@

Reef median depth had a moderate positive correlation with the number of years a reef
maintains coral cover above the assumed carbonate budget threshold required to sustain live
coral cover ($\theta_{0}$). Greater median depth was a key characteristic defining reefs
that sustain coral cover above $\theta_{0}$ (@fig-access-esm1-5-depth-conn-carbonate;
@sfig-access-esm1-5-mean-carb-budget; @sfig-access-esm1-5-carb-budget-corr). Unsurprisingly,
the number of years above $\theta_0$ substantially decreased as $\theta_0$ is increased, as
does the Spearman's correlation (@sfig-access-esm1-5-carb-budget-corr). Decreasing
correlation levels are likely due to an increasing number of reefs being unable to maintain
coral cover above $\theta_0$ as it is increased, with the distribution of years above
threshold becoming increasingly right-skewed (@fig-access-esm1-5-depth-conn-carbonate). In
contrast, weighted incoming connectivity strength (log10 scale) was found to have no
correlation with the number of years a reef maintains cover above the required thresholds
(@sfig-access-esm1-5-carb-budget-corr).

![Number of years reef coral cover exceeds the assumed carbonate budget threshold in terms
of coral cover ($\theta_0$) between 2025 to 2075 (y-axis) under ACCESS-ESM1-5. Left panel
shows relationship between reef median depth, while the right panel shows the same for
log10-weighted incoming connectivity values. Point colour (purple to yellow) represents
($\theta_0$), with brighter shades indicating a higher $\theta_0$. Deeper reefs have
increased number of years above the considered range of $\theta_0$, while no relationship
between weighted incoming connectivity and carbonate budget maintenance was
found.](../figs/ACCESS-ESM1-5/depth_conn_carbonate_comparison.png){#fig-access-esm1-5-depth-conn-carbonate
fig-align="center" width="14.82cm"}

>>>>

Partial Dependence Analysis revealed reef median depth, carrying capacity and mean time
series DHW level have the highest feature importance scores, accounting for 29%, 15% and 15%
predictive importance respectively (@fig-ts-cluster-pdp; Panel A). Source-cover weighted
incoming connectivity, outgoing connectivity strength, source-to-sink connectivity ratio and
larval retention probability account for \~12%, 11%, 10% and 8% respectively.

@@

![Feature importance scores (Panel A) of the Random Forest classifier, and Partial
Dependence Plots (PDPs) for reef characteristics of interest indicating the probability of a
reef being classified to one of the three cover clusters (Panel B). PDPs indicate the
probability of a reef from the test dataset being clustered into each cover cluster
depending on the predictor value on the x axis, when adjusted for values of other
predictors. Green, orange and blue lines indicate cluster assignment probability for Low,
Medium and High clusters respectively. Transparent grey points above probability lines
represent predictor values present in the test dataset, providing a rough indication of
predictor value distribution across assessed
reefs.](../figs/ts_cluster_rf_feature_analysis.png){#fig-ts-cluster-pdp fig-align="center"
width="14.82cm"}

First-order PDA showed minimal influence of incoming, outgoing and source-to-sink ratio
connectivity metrics on cluster assignment probabilities. Two-way analyses revealed little
interaction between depth and incoming and outgoing connectivity metrics for reef clustering
probabilities (@fig-ts-cluster-pdp-twoway-incoming; @sfig-ts-cluster-pdp-twoway-outgoing).
Reefs with a median depth around 15m and high incoming connectivity may benefit slightly,
with higher probability of High cover cluster assignment compared to 15m - depth reefs with
lower connectivity.

@@

Individual feature analysis for carrying capacity indicates that its influence on cluster
assignment is concentrated at the low range of predictor values (@fig-ts-cluster-pdp). This
aligns with the right-skewed distribution of carrying capacity across reefs for these three
predictor variables. Beyond this low-end range, partial dependence becomes near constant,
indicating outlier values for these factors are not influential. Assessing random forest
prediction accuracy across bioregions, reveals increasing performance as mean bioregion reef
depth increases (@sfig-rf-predictor-perf). This highlights that unexplained variation in the
test dataset (33% of variation) may be concentrated in shallow bioregions where available
model predictors do not capture differences between clustered reefs.

@@

Reef median depth had a moderate positive correlation with the number of years a reef
maintains coral cover above the assumed carbonate budget threshold required to sustain live
coral cover ($\theta_{0}$). Greater median depth was a key characteristic defining reefs
that sustain coral cover above $\theta_{0}$ (@fig-access-esm1-5-depth-conn-carbonate;
@sfig-access-esm1-5-carb-budget-corr; @sfig-access-esm1-5-mean-carb-budget). Unsurprisingly,
the number of years above $\theta_0$ substantially decreased as $\theta_0$ is increased, as
does the Spearman's correlation (@sfig-access-esm1-5-carb-budget-corr). Decreasing
correlation levels are likely due to an increasing number of reefs being unable to maintain
coral cover above $\theta_0$ as it is increased, with the distribution of years above
threshold becoming increasingly right-skewed (@fig-access-esm1-5-depth-conn-carbonate). In
contrast, weighted incoming connectivity strength (log10 scale) was found to have no
correlation with the number of years a reef maintains cover above the required thresholds
(@sfig-access-esm1-5-carb-budget-corr).

![Number of years reef coral cover exceeds the assumed carbonate budget threshold in terms
of coral cover ($\theta_0$) between 2025 to 2075 (y-axis) under ACCESS-ESM1-5. Left panel
shows the relationship between reef median depth, while the right panel shows the same for
log10-weighted incoming connectivity. Point colour (purple to yellow) represents
($\theta_0$), with brighter shades indicating a higher $\theta_0$. Deeper reefs have an
increased number of years above the considered range of $\theta_0$, while no relationship
between weighted incoming connectivity and carbonate budget maintenance was
found.](../figs/ACCESS-ESM1-5/depth_conn_carbonate_comparison.png){#fig-access-esm1-5-depth-conn-carbonate
fig-align="center" width="14.82cm"}
